25 years of Social Work Research

### 2015-03-22 dataCleaning.Rmd
Added new front matter to manuscript.

barplot(diff(nsf$n.grads), ylab="Change in value", main="Growth")

@@

# Introduction

It is estimated that 90% of all data in human history has been created in the last two years -- roughly 2.5 quintillion bytes of data every day [IBM, nd-a](http://www-01.ibm.com/software/data/bigdata/what-is-big-data.html).  _Big data_ has become ubiquitious in the business community and has recently captured the attention of governmental organizations given its potential to generate actionable insights from a wide range of data sources [IBM, nd-b](http://www-01.ibm.com/software/data/bigdata/industry-government.html).  The concept is most commonly understood with respect to _volume_ -- that is, an amount of data that exceeds the storage capacity of conventional database systems.  However, _big data_ is also distinguished from traditional data with respect to _velocity_ and _variety_.  Velocity refers to data arriving at a rate that exceeds conventional computing power, and variety refers to data structures that are not directly amenable to analysis with conventional software packages.  Volume, velocity and variety are commonly referred to as the three _V's_.

Big data comes from

Although the popularity of In fact, the interest in big data is growing in popularity as an internet search term, while ``social work'' is experiencing a decline.

(but not limited to) sensors that capture events in real time, GPS, social media sites, digitized text, electronic audio and video, purchase transactions, and weblogs.  These data provide rich opportunities for social work research that are

>>>>

@@

### Data Revolution

It is estimated that 90% of all the data in human history has been generated in the past two years, and the amount of data continues to grow by roughly 2.5 quintillion bytes every day [(IBM, nd-a)](http://www-01.ibm.com/software/data/bigdata/what-is-big-data.html).  We have unprecedented opportunities for deriving actionable insights from these data to advance the human condition.  The significance continues to gain the attention of national and international organizations.  For example, the Independent Expert Advisory Group (IEAG) of the United Nations prepared a report entitled, _A World That Counts:  Mobilising the Data Revolution for Sustainable Development_ [(IEAG, 2014)](http://www.undatarevolution.org/report/).  The IEAG (2014) argues that "Data are the lifeblood of decision-making and the raw material for accountability.  Without high-quality data providing the right information on the right things at the right time; designing, monitoring and evaluating effective policies becomes almost impossible" (p. 2).  The United States federal government established an open data initiative, motivated by the awareness that data can have the greatest impact when it is "accessible, discoverable, and usable" [(www.data.gov)](www.data.gov).  Data from all levels of government that is made completely open can lead to "cost savings, efficiency, fuel for business, improved civic services, informed policy, performance planning, research and scientific discoveries, transparency and accountability, and increased public participation in the democratic dialogue" (www.data.gov).  At the time of preparing this article, over 124,000 data sets were freely available through www.data.gov among a diverse set of topic areas:  agriculture, business, climate, consumers, ecosystems, education, energy, finance, health, local government, manufacturing, ocean studies, public safety, and science and research.

In 2012, the NIH established the Big Data to Knowledge (BD2K) initiative, recognizing the potential of data to advance our understanding of human health and disease.  This initiative was designed to support four broad goals:  1) to facilitate biomedical digital assets by making them discoverable, accessible, and citable; 2) to conduct research and development methods, software and tools to analyze biomedical big data; 3) to enhance training in the development of use and tools necessary for biomedical big data science; and 4) to support a data ecosystem that accelerates discovery as part of the digital enterprise (NIH).  At roughly the same time, the National Science Foundation (NSF) created new initiatives to harness the potential of the data revolution, arguing that ”Data are motivating a profound transformation in the culture and conduct of scientific research in every field of science and engineering ... American scientists must rise to the challenges and seize the opportunities afforded by this new, data-driven revolution" (Suresh, XXX).  The programs of the NSF are organized around capacity building and innovative applications.  Capacity building activities focus on developing fundamental theories, techniques, methodologies and technologies that are broadly applicable to big data problems.  Innovative applications relate to specific domains advancing innovative ideas that provide solutions “with potential for a broader impact on data science and its applications.”

### Data Science and Social Work

We are, indeed, living in a data revolution.  The relevance to social work research is clear — we have the opportunity to study human behavior and other social phenomena in real time, with vastly improved reliability in our measurements, and at scale.  Data comes from a wide range of sources, including (but not limited to) social media interactions, electronic medical records, audio and video recordings, click stream data, transaction data, web logs, embedded sensors, and GPS.  New tools allow blending of data sets, unique ways to analyze data, and methods of visualizations that can produce insights limited only by imagination.  As stated by the NIH, "lack of appropriate tools, poor data accessibility, and insufficient training, are major impediments to rapid translational impact" in this new era of data.  It is essential that the field of social work fully embraces the data revolution and considers the extent to which the existing research infrastructure and training is preparing both social work researchers and providers to be effective and efficient in a data rich world.  The potential of the data revolution is reflected in the current _Grand Challenge_ entitled, “Harness technology for social good.”  However, we believe that it the data revolution goes beyond this single _Grand Challenge_ as it provides the opportunity to be informative in all areas of social work research.

Data science serves as a useful framework for broadening our existence research infrastructure and training opportunities.  Data science is similar to traditional academic research, as it includes the key roles of substantive expertise and traditional research methodologies.  Much of the theoretical basis comes from the field of statistics (Zumel & Mount, 2014), but the famous statistician William Cleveland argued that it is an interdisciplinary field that is much larger than statistics (Cleveland, XXX).  A major difference is the data science draws heavily from knowledge the areas of computer science, software engineering, and information technology.  Unlike traditional academic research, data science is particularly in tackling the _3 V’s_ of big data:  _volume_, _variety_, and _velocity_.  More specifically, in data-rich environments, researchers encounter an amount or volume of data that often exceeds the capacity of traditional database systems.  Data also come in a variety of formats, ranging from highly structured data contained in relational databases to massive stores of unstructured text-data culled from social media.  Many data opportunities go beyond the traditional approach of downloading a dataset and analyzing it with an off-the-shelf statistical package like SPSS, Stata, or SAS.  Now, data can be accessed in real time through application program interfaces (API), which brings along the challenge of _velocity_ — that is, data can be transmitted at a rate that grossly exceeds the computing power of desktop computers.  These are practical challenges that need to be addressed in order to maximize the potential of these new data.

Another feature that distinguishes data science from traditional academic research is the focus on generating _actionable insights_ from data rather than theory building and hypothesis testing. In fact, data science is more interested in deriving useful insights from correlational patterns as opposed to uncovering causal mechanisms.  The focus is on the development data products (Davenport, XXX), as opposed to reporting analysis.  Data products are tools or services that generate actionable insights from the data itself, including (but not limited to) automated reports or dashboards, recommendation systems, prediction algorithms, decision tools, and interactive visualizations that help people with non-technical background make sense of complex data.  Data science also values the philosophy of _openness_ or open science.  Thus, much effort is devoted to creating sustainable systems that make data as widely and freely accessible as possible, and developing and promoting the use of open source tools for managing and analyzing data.  The concept of reproducible research is a _rule_ rather than the _exception_.  This involves making both data and statistical code available alongside research reports.  This helps ensure the quality of the research, in addition to allowing researchers to make advances without having duplicate work that has already been done.  Data science is by no means incompatible with traditional academic research.  We believe that the tools, research strategies, and principles of data science are complementary to traditional ways of knowledge building in social work, offering many unique opportunities to improve the quality, efficiency, and impact of this area of research.

### Overview of Current Study

In the current study, we seek to the opportunities of the data revolution and data science by applying a range of tools and strategies to construct an historical base of social work research.  The data are article records harvested from existing bibliographic databases.  Article records are already widely used in various bibliometric studies, particularly those that are intended to reveal publication networks within scientific communities.  It does so by creating links between co-authors, and then visually representing the edges and vertices of the network.  This is also done with article citation histories, to show influence over time.  In this study, we were interested in the various article meta-data that is common to many scientific article records.  For example, article records a single article record in psycInfo can have as many as XX pieces of meta-data including:  dates, journal titles, article titles, author names, author affiliations, full text abstracts, keywords, subject classification, methodology, location, populations, page numbers, source of funding.  This is a rather large amount of information about a single paper, at least compared to the amount of information contained in article records before the electronic records.  Large collections of article records may contain potentially interesting trends in the research, assuming the selection of the article records were conducted in a way that allowed meaningful inferences to be derived.

In the current study, we sought to capture every journal article record published in every social work journals over the in past quarter century (1989-2012) and to explore the various meta-data to uncover potentially interesting trends in the social work scholarship.  For example, we can reasonably infer estimates of _size_ by counting the number of unique article records.  And, growth of the field can be inferred by examining the number of article records over time.  Similarly, we examine the extent to which _team science_ has become part of the research practice, as this has implications for training of social work researchers.  We also use location meta-data to understand the extent to which different areas of the world have and have not been the focus of social work research.  Finally, we used topic modeling to extract topic areas that define the focus of social work research.  The execution of this research was grounded in the tools and values of data science.  Thus, all data were managed and analyzed using open source software, R.  We also used an open source authoring system that integrates the statistical analysis with the text of the manuscript.  Along with this manuscript, the authors will also release the actual data and statistical code, thereby meeting the requirements of _reproducible research_, as defined by King ().  To our knowledge, this is the first time fully reproducible research (as defined by King, XXXX), has been published in a social work journal.

Davenport: http://blogs.wsj.com/cio/2014/06/25/so-you-want-to-build-a-data-product/

http://www.nsf.gov/pubs/2015/nsf15544/nsf15544.htm

http://www.nsf.gov/news/news_summ.jsp?cntn_id=123607&org=NSF&from=news
